Wait-free Queues with Polylogarithmic Step Complexity

### 2022-10-06 thesis/thesis.tex
Finalize proof, apply Oct 5 notes

@@

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{: \# enqueue, dequeue operations in the prefix for the block}

\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}
  \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}
  \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}
  \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}
  \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
\end{itemize}

@@

\it{Abbreviations:}
\begin{itemize}
 \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
 \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\})}
\end{itemize}
\end{footnotesize}

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the the response to the $D_{root,b,i}$.}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \Return \tt{null} \label{returnNull}
\Else
\State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
\cmt{$E_e(root)$ is the response.}
\State \Return \tt{root.GetEnqueue(root.\Call{DSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
\State \tt{start= end-1} \label{dsearchStart}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile \label{dsearchEnd}
\State \tt{b= root.BinarySearch(sum\sub{enq}, e, start, end)} \label{dsearchBinarySearch}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
\State \tt{blocks[\head]= blk} 
\State \tt{\head+=1} \label{appendEnd} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

@@

\Function{void}{Advance}{\sl{int} h}
\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
\State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
\State \tt{head.CAS(h, h+1)} \label{incrementHead}
\EndFunction{Advance}

\pagebreak
\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
\Function{int}{BinarySearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}

\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}. Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
%\State \Return \tt{result block's index}
\EndFunction{BinarySearch}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
\EndFor
\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
\EndIf

\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
\EndFunction{CreateBlock}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_i(this, b)=D_y(root,x)$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= parent.left==n ? left: right} 
\State \tt{sb= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ?  blocks[b].super: blocks[b].super+1}\label{computeSuper}

\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-left}} \label{considerPreviousLeft}
\EndIf

\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
\State \tt{i+= parent.blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{sb, i}
\EndIf
\EndFunction{IndexDequeue}

@@

\begin{definition}[Linearization] \label{def::lin}
 $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
\end{definition}

\begin{observation}\label{sumToNum}
$n.\nf{blocks[}i\nf{].sum\sub{x}}-n.\nf{blocks[}j\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$ ($i<j$)
\end{observation}

\begin{lemma}\label{lem::numX}
Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
\begin{enumerate}[label=(\arabic*)]
    \item If $n$ is an internal node $B\nf{.num\sub{x-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
    \item If $n$ is an internal node $B\nf{.num\sub{x-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
    \item $B\nf{.num\sub{x}}=\Big|E(n,b)\Big|$
\end{enumerate}
\end{lemma}
\begin{proof}
 We prove the claim by induction on height of node $n$. Base case (3) for leaves is trivial. Supposing the claim is true for $n$'s children, we prove the correctness of claim for $n$.
\begin{align*}
    B\nf{.num\sub{x-left}}=&B.\nf{sum\sub{x-left}}-B^\prime.\nf{sum\sub{x-left}}& \text{Definition of }\nf{num\sub{x}}\\
    =&B^\prime.\nf{sum\sub{x-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}-B^\prime.\nf{sum\sub{x-left}} & \nf{CreateBlock}\\
    =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}\\
    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{x}} &\text{Observation} \ref{sumToNum}
\end{align*}
The last line holds because of the induction hypothesis. (2) is similar to (1). Now we prove (3) starting from the Definition of $E(n,b)$.
\begin{align*} 
E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{x-left}}+B\nf{.num\sub{x-right}}=B\nf{.num\sub{x}}$
\end{proof}

\begin{corollary}\label{lem::sumX}
Let $B$ be $n\nf{.blocks[}b\nf{]}$ and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
\begin{enumerate}[label=(\arabic*)]
    \item If $n$ is an internal node $B\nf{.sum\sub{x-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
    \item If $n$ is an internal node $B\nf{.sum\sub{x-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
    \item $B\nf{.sum\sub{x}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
\end{enumerate}
\end{corollary}

@@

\begin{lemma}\label{blockSize}
Each block contains at most one operation of each process, and therefore at most $p$ operations in total.
\end{lemma}
\begin{proof}
To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $b$ in node $n$. Without loss of generality $op_1$ is invoked earlier than $op_2$. Process $p$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins. By Corollary \ref{lem::appendExactlyOnce} before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root. Since $b$ contains $op_2$, it must be created after $op_2$ is invoked. This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$. The existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
\end{proof}

@@

\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} so that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
\begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
If $\nf{root.blocks[$end$]}\neq \nf{null}$ and $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
\nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$. From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order. Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} is also non-decreasing. Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumX}. Block \nf{root.blocks[$b$]} contains $E_i(root,b)$. The doubling search on Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$. In Line \ref{dsearchBinarySearch}, the binary search finds the $b$ required in the range mentioned. Finally $i$, is computed from the definition of \nf{sum\sub{enq}} and lemma \ref{lem::sumX}.
\end{proof}

\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of node \nf{n}. For the base case, suppose $n$ is a leaf. Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf. Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.

>>>>

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}
  \textsf{: \# enqueues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}
  \textsf{: \# dequeues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}
  \textsf{: \# enqueues in \tt{right.blocks[1..end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}
  \textsf{: \# dequeues in \tt{right.blocks[1..end\sub{right}]}}
\end{itemize}

@@

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{: \# enqueue, dequeue operations in the prefix for the block}
\end{itemize}

@@

\it{Abbreviations used in the code and the proof of correctness.}
\begin{itemize}
 \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{ enq, deq, enq-left, enq-right, deq-left, deq-right\})}
\end{itemize}
\end{footnotesize}

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the response to $D_{root,b,i}$.}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \Return \tt{null} \label{returnNull}
\Else
\State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
\cmt{$E_e(root)$ is the response.}
\State \Return \tt{root.GetEnqueue(root.\Call{DSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$.}
\State \tt{start= end-1} \label{dsearchStart}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile \label{dsearchEnd}
\State \tt{b= root.BinarySearch(sum\sub{enq}, e, start, end)} \label{dsearchBinarySearch}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\Function{void}{Append}{\sl{block} blk} \cmt{Only called by the owner of the leaf.}
\State \tt{blocks[\head]= blk} 
\State \tt{\head+=1} \label{appendEnd} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

@@

\Function{void}{Advance}{\sl{int} h}
\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
\State \tt{blocks[h].super.CAS(null, h\sub{p}} \label{setSuper1}
\State \tt{head.CAS(h, h+1)} \label{incrementHead}
\EndFunction{Advance}

\pagebreak
\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} greater than or equal to \tt{i}}
\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{i} of the given prefix sum \tt{field}. Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.
\Function{int}{BinarySearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
\State \Return \tt{min\{j: blocks[j].f$\geq$i\}}
\EndFunction{BinarySearch}

@@

\Statex\cmt{Creates and returns the block to be installed in \tt{blocks[i]}. Created block includes \tt{left.blocks[index\sub{prev}+1..index\sub{last}]} and \tt{right.blocks[index\sub{prev}+1..index\sub{last}]}.}
\Function{Block}{CreateBlock}{\sl{int} i} 
\State \tt{block new= \Call{new}{\sl{block}}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{new.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + }
\Statex \hspace{10.7em}\tt{block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + }
\Statex \hspace{10.7em}\tt{block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
\EndFor
\If{\tt{this} \bf{is} \tt{root}}
\State \tt{new.\size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}}
\Statex \hspace{8.5em} \tt{- new.num\sub{deq}, 0)}\label{computeLength}
\EndIf

\State \Return \tt{new}
\EndFunction{CreateBlock}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_i(this, b)=D_y(root,x)$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n ? left: right)} 
\State \tt{sb= (parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ?  blocks[b].super: blocks[b].super+1)}\label{computeSuper}

\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-left}} \label{considerPreviousLeft}
\Else \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
\State \tt{i+= parent.blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{sb, i}
\EndIf
\EndFunction{IndexDequeue}

@@

The linearization ordering is given by the order that operations appear in the blocks in the root.
\begin{definition}[Linearization] \label{def::lin}
 $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
\end{definition}

\begin{observation}\label{sumToNum}
For any node $n$ and indices $i<j$ of \nf{blocks} in $in$, we have
$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where \nf{x} in $\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
\end{observation}

Next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
\begin{lemma}\label{lem::numX}
Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively.
\begin{enumerate}[label=(\arabic*)]
    \item If $n$ is an internal node $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
    \item If $n$ is an internal node $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
\end{enumerate}
\end{lemma}
\begin{proof}
 We prove the claim by induction on height of node $n$. Base case (3) for leaves is trivial. Supposing the claim is true for $n$'s children, we prove the correctness of the claim for $n$.
\begin{align*}
    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}& \text{Definition of }\nf{num\sub{enq}}\\
    =& B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}} & \nf{CreateBlock}\\
    =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
    =& \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} &\text{Observation } \ref{sumToNum}\\
    =& \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big| & \textrm{Induction hypothesis (3)}
\end{align*}
The last line holds because of the induction hypothesis (3). (2) is similar to (1). Now we prove (3) starting from the Definition of $E(n,b)$.
\begin{align*} 
E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
\end{proof}

Next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
\begin{corollary}\label{lem::sumX}
Let $B$ be $n\nf{.blocks[}b\nf{]}$ and \nf{enq} be in $\{\nf{enq},\nf{deq}\}$.
\begin{enumerate}[label=(\arabic*)]
    \item If $n$ is an internal node $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
    \item If $n$ is an internal node $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
\end{enumerate}
\end{corollary}

@@

\begin{lemma}\label{blockSize}
Each block contains at most one operation of each process, and therefore at most $p$ operations in total.
\end{lemma}
\begin{proof}
To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $b$ in node $n$. Without loss of generality $op_1$ is invoked earlier than $op_2$. Process $p$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins. By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root. Since $b$ contains $op_2$, it must be created after $op_2$ is invoked. This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$. The existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
\end{proof}

@@

\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
\begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
If $\nf{root.blocks[$end$]}\neq \nf{null}$ and $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
 From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order. Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing. Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumX}. Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.  Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$. Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blocks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}. By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$. Finally $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
\end{proof}

\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of node $n$. For the base case, suppose $n$ is a leaf. Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf. Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
